---
title: "3125 Biostatistics Module 5"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
    Tutorials for 3125 Biostatistics. Adapted from https://daviddalpiaz.github.io/appliedstats/.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(Biostatistics)
library(Lahman)
library(ggplot2)
library(binom)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
```

## Learning Outcomes 

This module, we will focus on studying the inferences for the proportions (categorical variables). After reading this tutorial you will be able to:

 - Install an R package.
 - Sampling distribution of a proportion by repeated sampling from a known population.
 - Inference for one population proportion.
 - Inference for two population proportions.
 
## Binomial distribution

#### Load required packages
We’ll use the binom package to calculate a confidence interval for a proportion using the Agresti-Coull method. They need to be loaded only once per session.

You might need to install the binom package on your computer if you have not already done so.

```{r, eval=FALSE}
install.packages("binom", dependencies = TRUE) # only if not yet installed
library(binom)
```

#### Calculating binomial probabilities:

Calculate a binomial probability, the probability of obtaining $X$ successes in $n$ trials when trials are independent and probability of success p is the same for every trial.

#### Binomial probability
Suppose we have 27 flowers, we are interested in the probability of the number of left-handed flowers.
So, the probability of getting exactly 6 left-handed flowers when n=27 and p=0.25 is

```{r}
dbinom(6, size = 27, prob = 0.25)
```

#### Table of probabilities
Tabulate the probability distribution of all possible values for the number of left-handed flowers out of 27.

```{r, echo=FALSE}
xsuccesses <- 0:27
probx <- dbinom(xsuccesses, size = 27, prob = 0.25) 
probTable <- data.frame(xsuccesses, probx)
probTable
```

#### Histogram of probability distribution
The histogram of this binomial distribution for the number of left-handed flowers when n=27 and p=0.25.

```{r, echo=FALSE}
ggplot(probTable, aes(x = xsuccesses, y = probx)) + 
    geom_bar(stat = "identity", fill = "firebrick", col = "black") +
    labs(x = "Number of left-handed flowers", y = "Probability") +
    theme_classic()
```


## Sampling distribution of a proportion
Compare sampling distributions for the proportion based on n=10 and n=100.

Repeat taking a random sample size of 10 observations (n=10) from a population having probability of success p=0.25 for 10000 times. Convert to proportions by dividing by the sample size. Visualize the distribution of all the 10000 proportions in a histogram.

```{r, echo=FALSE}
successes10 <- rbinom(10000, size = 10, prob = 0.25)
proportion10 <- successes10 / 10
ggplot(data = data.frame(proportion10), aes(x = proportion10)) + 
    stat_bin(aes(y=..count../sum(..count..)), fill = "firebrick", col = "black",
             boundary = 0, closed = "left", binwidth = 0.1) +
    labs(x = "Sample proportion", y = "Probability") + 
    coord_cartesian(xlim = c(0, 1)) +
    theme_classic()
```

Repeat taking a random sample size of 100 observations (n=100) from a population having probability of success p=0.25 for 10000 times. Convert to proportions by dividing by the sample size. Visualize the distribution of all the 10000 proportions in a histogram.

```{r, echo=FALSE}
successes100 <- rbinom(10000, size = 100, prob = 0.25)
proportion100 <- successes100 / 100
ggplot(data = data.frame(proportion100), aes(x = proportion100)) + 
    stat_bin(aes(y=..count../sum(..count..)), fill = "firebrick", col = "black",
             boundary = 0, closed = "left", binwidth = 0.01) +
    labs(x = "Sample proportion", y = "Probability") + 
    coord_cartesian(xlim = c(0, 1)) +
    theme_classic()
```


## Inference for one population proportion

In this tutorial, we will talk about how to make inference for one population proportion. How to calculate the Standard error and 95% confidence interval for a proportion using the Agresti-Coull method for the confidence interval.

We will use a example data with gender information for a group of sea turtle. First, we read and inspect data
```{r}
turtles <- read.csv(url("https://raw.githubusercontent.com/xzhang47/3125/main/chap07e3GreenTurtleSex.csv"), stringsAsFactors = FALSE)
head(turtles)
```

Here are the null and alternative hypotheses:

H0: The proportion of female sea turtle is 0.75. 
HA: The proportion of female sea turtle is not 0.75. 


Frequency table of female and male offspring number.

```{r}
turtleTable <- table(turtles$Sex)
turtleTable
```

Calculate proportion
Estimate the proportion of offspring that are female.

```{r}
n <- sum(turtleTable)
n
```

```{r}
X <- turtleTable[1]
pHat <-  X / n
pHat
```

Standard error of proportion
```{r}
sqrt( (pHat * (1 - pHat))/n )
```

Second, use the binom package. The result will be very slightly different from the one you calculated above because the formula we use in the book takes a slight shortcut.

```{r}
binom.test(X, n = n, p = 0.75)
```

## Inference for two population proportions

Here are the null and alternative hypotheses:

H0: There is no difference of the proportion of divorce after cancer diagnosis between male and female. 
HA: There is a difference of the proportion of divorce after cancer diagnosis between male and female.

In this tutorial, we will learn how to look at, and test for, the difference of proportions of a categorical variable between two populations. We’ll again use the a `divorce` data set as an example, looking at the differences of proportions of divorces between gender groups after cancer diagnosis. Let’s load the data from the .csv file (Gender disparity in the rate of partner abandonment in patients with serious medical illness):

```{r}
divorce <- read.csv(url("https://raw.githubusercontent.com/xzhang47/3125/main/chap09q03DiseaseAndDivorce.csv"), header = TRUE)
divorce
```

We first can acquire the contingency table using `table()` and saved it as an object `divorce_table`:
```{r}
divorce_table <- table(divorce$illPartner, divorce$divorce)
```

Then using the saved object `divorce_table` in the function `prop.test()`, we can get the output for a 2-sample test for equality of proportions without continuity correction as:
```{r}
prop.test(divorce_table, correct=FALSE) 
```

As a result, we found the p-value and confidence interval from the output directly. Or, we can get the p-value separately:

```{r}
prop.test(divorce_table, correct=FALSE)$p.value 
```
The p-value is $`r prop.test(divorce_table, correct=FALSE)$p.value`$ smaller than 0.05, we reject the null hypothesis and conclude there is statistical difference of divorce rate after cancer diagnosis between gender groups.
95% Confidence interval:
```{r}
prop.test(divorce_table, correct=FALSE)$conf.int 
```

The 95% confidence interval estimate of the difference between the female proportion of Aboriginal students and the female proportion of Non-Aboriginal students is between -2.7% and 20.9%.

