---
title: "3125 Biostatistics Module 1"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
    Tutorials for 3125 Biostatistics. Adapted from https://daviddalpiaz.github.io/appliedstats/.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(Biostatistics)
library(Lahman)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
```

## Learning Outcomes 

This module, we will introduction and analysis of a single variable. After reading this tutorial you will be able to:

- Basics of using R.
- Visualizing and describing a single categorical variable.
- Visualizing and describing a single quantitative variable.

## Introduction to `R`

To get started, we'll use `R` like a simple calculator.

### Basic Calculations

#### Addition, Subtraction, Multiplication and Division {-}

| Math          | `R`     | Result    |
|---------------|---------|-----------|
| $3 + 2$       | `3 + 2` | `r 3 + 2` |
| $3 - 2$       | `3 - 2` | `r 3 - 2` |
| $3 \cdot2$    | `3 * 2` | `r 3 * 2` |
| $3 / 2$       | `3 / 2` | `r 3 / 2` |

#### Exponents  {-}

| Math         | `R`             | Result            |
|--------------|-----------------|-------------------|
| $3^2$        | `3 ^ 2`         | `r 3 ^ 2`         |
| $2^{(-3)}$   | `2 ^ (-3)`      | `r 2 ^ (-3)`      |
| $100^{1/2}$  | `100 ^ (1 / 2)` | `r 100 ^ (1 / 2)` |
| $\sqrt{100}$ | `sqrt(100)`     | `r sqrt(100)`     |

#### Mathematical Constants  {-}

| Math         | `R`             | Result            |
|--------------|-----------------|-------------------|
| $\pi$        | `pi`            | `r pi`            |
| $e$          | `exp(1)`        | `r exp(1)`        |

#### Logarithms  {-}

Note that we will use $\ln$ and $\log$ interchangeably to mean the natural logarithm. There is no `ln()` in `R`, instead it uses `log()` to mean the natural logarithm.

| Math              | `R`                 | Result                |
|-------------------|---------------------|-----------------------|
| $\log(e)$         | `log(exp(1))`       | `r log(exp(1))`       |
| $\log_{10}(1000)$ | `log10(1000)`       | `r log10(1000)`       |
| $\log_{2}(8)$     | `log2(8)`           | `r log2(8)`           |
| $\log_{4}(16)$    | `log(16, base = 4)` | `r log(16, base = 4)` |

#### Trigonometry  {-}

| Math            | `R`           | Result          |
|-----------------|---------------|-----------------|
| $\sin(\pi / 2)$ | `sin(pi / 2)` | `r sin(pi / 2)` |
| $\cos(0)$       | `cos(0)`      | `r cos(0)`      |

### Functions

Throughout the tutorial we will be using functions. Here we will discuss some of their details.

```{r, eval = FALSE}
function_name(arg1 = 10, arg2 = 20)
```

To use a function, you simply type its name, followed by an open parenthesis, then specify values of its arguments, then finish with a closing parenthesis. 

An **argument** is a variable which is used in the body of the function. Specifying the values of the arguments is essentially providing the inputs to the function.

### Getting Help

In using `R` as a calculator, we have seen a number of functions: `sqrt()`, `exp()`, `log()` and `sin()`. To get documentation about a function in `R`, simply put a question mark in front of the function name and RStudio will display the documentation, for example: 

```{r, eval = FALSE}
?log
?sin
?paste
?lm
```


## Dataset Setup

### Data Variable Types

In this course, we focused on analyzing two types of variables: 

| **Categorical (Qualitative)** | **Quantitative (Measurement)**    |
|-------------------------------|-----------------------------------|
| Values measured in groups     | Values measured in numeric        |
| or categories                 | values with meaning (units/counts)|
| Examples:                     |   Examples:                       |
| Gender                        |   # of minutes                    |
| Favorite color                |   weight (in pounds)              |
| Type of car                   |   snowfall (in inches)            |   


In `R`, the data can be stored in various number of basic data *types*.

- Numeric
    - Also known as Double. The default type when dealing with numbers.
    - Examples: `1`, `1.0`, `42.5`
- Integer
    - Examples: `1L`, `2L`, `42L`
- Complex
    - Example: `4 + 2i`
- Logical
    - Two possible values: `TRUE` and `FALSE`
    - You can also use `T` and `F`, but this is *not* recommended.
    - `NA` is also considered logical.
- Character
    - Examples: `"a"`, `"Statistics"`, `"1 plus 2."`

Among them, numeric or integer is often used to record the quantitative variables while character is often used to store the categorical variables. Categorical variables can also be recorded as numeric in `R`. Complex and logical types won't be commonly used in this course.

### Data Structures

`R` also has a number of basic data *structures*. A data structure is either homogeneous (all elements are of the same data type) or heterogeneous (elements can be of more than one data type).

| Dimension | **Homogeneous** | **Heterogeneous** |
|-----------|-----------------|-------------------|
| 1         | Vector          | List              |
| 2         | Matrix          | Data Frame        |
| 3+        | Array           |                   |

### Vectors

Many operations in `R` make heavy use of **vectors**. Vectors in `R` are indexed starting at `1`. That is what the `[1]` in the output is indicating, that the first element of the row being displayed is the first element of the vector. Larger vectors will start additional rows with `[*]` where `*` is the index of the first element of the row.

Possibly the most common way to create a vector in `R` is using the `c()` function, which is short for "combine."" As the name suggests, it combines a list of elements separated by commas. 

```{r}
c(1, 3, 5, 7, 8, 9)
```

Here `R` simply outputs this vector. If we would like to store this vector in a **variable** we can do so with the **assignment** operator `=`. In this case the variable `x` now holds the vector we just created, and we can access the vector by typing `x`.

```{r}
x = c(1, 3, 5, 7, 8, 9)
x
```

### Data Frames

For this course, we will be using **data frame** most often which will be the most common way that we store and interact with data.

#### Create a dataset using vectors
The first method to create a data frame is using the function `data.frame` to combine multiple **vector** into the dataframe.

```{r}
example_data = data.frame(x = c(1, 3, 5, 7, 9, 1, 3, 5, 7, 9),
                          y = c(rep("Hello", 9), "Goodbye"),
                          z = rep(c(TRUE, FALSE), 5))
```

A data frame is a **list** of vectors. So, each vector must contain the same data type, but the different vectors can store different data types. 

```{r}
example_data
```

#### Import dataset stored in R packages

First we will use a dataset supplied by and R package `Biostatistics`. Earlier we looked at installing packages, in particular the `Biostatistics` package. (This package contains several biostatistics datasets which will be used for demonstration). 

```{r, message = FALSE, warning = FALSE}
library(Biostatistics)
```

Inside the `Biostatistics` package, there is a dataset called `height_immunity`. By loading the package using the `library()` function, we can now access `height_immunity`.

When using data from inside a package, there are three things we would generally like to do:

- Look at the raw data.
- Understand the data. (Where did it come from? What are the variables? Etc.)
- Visualize the data.

To look at the data, we have two useful commands: `head()` and `str()`.

```{r}
head(height_immunity, n = 10)
```

The function `head()` will display the first `n` observations of the data frame. If not using the function, write the name of dataset will print all observations.

```{r}
height_immunity
```

The function `str()` will display the "structure" of the data frame. It will display the number of **observations** and **variables**, list the variables, give the type of each variable, and show some elements of each variable. This information can also be found in the "Environment" window in RStudio.

```{r}
str(height_immunity)
```

As you can see, data frames have observations and variables. When displayed in the console or viewer, each row is an observation and each column is a variable. 

To understand more about the data set, we use the `?` operator to pull up the documentation for the data.

```{r, eval = FALSE}
?height_immunity
```

`R` has a number of functions for quickly working with and extracting basic information from data frames. To quickly obtain a vector of the variable names, we use the `names()` function.

```{r}
names(height_immunity)
```

To access one of the variables **as a vector**, we use the `$` operator.

```{r}
height_immunity$sex
height_immunity$age
```

We can use the `dim()`, `nrow()` and `ncol()` functions to obtain information about the dimension of the data frame.

```{r}
dim(height_immunity)
nrow(height_immunity)
ncol(height_immunity)
```

Here `nrow()` is also the number of observations, which in most cases is the *sample size*. `ncol()` is the number of variables.

Subsetting data frames can work much like subsetting matrices using square brackets, `[,]`. Here, we find fuel efficient vehicles earning over 35 miles per gallon and only display `manufacturer`, `model` and `year`.

```{r}
height_immunity[height_immunity$age > 35, c("age", "sex", "body.height")]
```

An alternative would be to use the `subset()` function, which has a much more readable syntax.

```{r, eval = FALSE}
subset(height_immunity, subset = age > 35, select = c("age", "sex", "body.height"))
```

Both approaches produce the same results. Which you use will be largely based on a given situation as well as user preference.

When subsetting a data frame, be aware of what is being returned, as sometimes it may be a vector instead of a data frame. 

#### Import dataset stored in R packages

Aternatively, we can also import data from various file types in into `R`.

First, write the dataset to your local computer hard disk by update the directory path: `data/height_immunity.csv` to your own computer path. This particular line of code assumes that the file `height_immunity.csv` exists in a folder called `data` in your current working directory.

```{r writecsv, exercise = TRUE, exercise.eval = FALSE}
write.csv(height_immunity, "data/height_immunity.csv", row.names = FALSE)
```

```{r summarise-check}
"Good job!"
```

To read this data into `R`, we would use the `read.csv()` function. Please update the directory path: "data/height_immunity.csv" to your own computer path.

```{r readcsv, exercise = TRUE, exercise.eval = FALSE}
height_immunity_data_from_csv = read.csv("data/height_immunity.csv")
```

Now you can use the function `head()` and other functions we introduced as above to explore the basic features of the imported dataset: `height_immunity_data_from_csv`.

```{r}
head(height_immunity_data_from_csv, n = 10)
```

## Visualizing and describing a single categorical variable
Now that we have some data to work with, and we have learned about the data at the most basic level, our next tasks is to visualize and describe/summarize the data. `R` has built in functions for a large number of summary statistics. 

### Categorical summary statistics 

For categorical variables, counts and percentages can be used for summary.

```{r}
table(height_immunity$sex)
table(height_immunity$sex) / nrow(height_immunity)
```

### Plotting

Next, a proper visualization can illuminate features of the data that can inform further analysis. We will look at two methods of visualizing data that we will use throughout the course:

- Barplots
- Pie Charts

#### Barplots

A barplot can provide a visual summary of a categorical variable, or a numeric variable with a finite number of values, like a ranking from 1 to 10.

```{r}
barplot(table(height_immunity$sex))
```

```{r}
barplot(table(height_immunity$sex),
        xlab   = "Gender",
        ylab   = "Frequency",
        main   = "Gender",
        col    = "dodgerblue",
        border = "darkorange")
```

#### Pie Charts

```{r}
pie(table(height_immunity$sex))
```

#### Exercise 1

In this exercise, use a dataset `quolls` in the R package `Biostatistics`, shown as below:

```{r}
head(quolls, n = 10)
```
    
```{r module1ex1, exercise = TRUE}
    
```

```{r module1ex1-solution}

```
    
<div id="module1ex1-hint">
**Hint:** Use`barplot()` to get the plot. 
</div>

```{r summariseex4-check}
"Great work!"
```

## Visualizing and describing a single quantitative variable

For numeric variables, we can summarize data with the center and spread. We'll again look at the `height_immunity` dataset from the `Biostatistics` package.

#### Central Tendency {-}

| Measure | `R`               | Result              |
|---------|-------------------|---------------------|
| Mean    | `mean(height_immunity$age)`   | `r mean(height_immunity$age)`   |
| Median  | `median(height_immunity$age)` | `r median(height_immunity$age)` |

#### Spread {-}

| Measure            | `R`              | Result             |
|--------------------|------------------|--------------------|
| Variance           | `var(height_immunity$age)`   | `r var(height_immunity$age)`   |
| Standard Deviation | `sd(height_immunity$age)`    | `r sd(height_immunity$age)`    |
| IQR                | `IQR(height_immunity$age)`   | `r IQR(height_immunity$age)`   |
| Minimum            | `min(height_immunity$age)`   | `r min(height_immunity$age)`   |
| Maximum            | `max(height_immunity$age)`   | `r max(height_immunity$age)`   |
| Range              | `range(height_immunity$age)` | `r range(height_immunity$age)` |


### Plotting

#### Histograms

When visualizing a single numerical variable, a **histogram** will be our go-to tool, which can be created in `R` using the `hist()` function.

```{r}
hist(height_immunity$age)
```

The histogram function has a number of parameters which can be changed to make our plot look much nicer. Use the `?` operator to read the documentation for the `hist()` to see a full list of these parameters.

```{r}
hist(height_immunity$age,
     xlab   = "Age in years",
     main   = "Histogram of Age",
     breaks = 12,
     col    = "dodgerblue",
     border = "darkorange")
```

Importantly, you should always be sure to label your axes and give the plot a title. The argument `breaks` is specific to `hist()`. Entering an integer will give a suggestion to `R` for how many bars to use for the histogram. By default `R` will attempt to intelligently guess a good number of `breaks`, but as we can see here, it is sometimes useful to modify this yourself.

#### Exercise 2

Use the data `quolls` from the R package Biostatistics, draw a histogram for the body length of quolls.
    
```{r module1ex2, exercise = TRUE}
    
```

```{r module1ex2-solution}
hist(quolls$bodylength)
```
    
<div id="module1ex2-hint">
**Hint:** Use `hist()` to generate a histogram.
</div>

```{r summariseex5-check}
"Great work! That was tough. Be sure you understand the function involved."
```

## Normal Distribution and Empirical Rules

As we have showed the histograms of the above quantitative variables, you may notice the shape of the histogram naturally followed a symmetric unimodal bell shape, bulging in the middle and tapering off at the ends. One most commonly known bell shape distribution is the normal distribution.

### The normal distribution

#### Characteristics of the normal distribution
- Symmetric around its mean
- Mean = median = mode
- The area under the normal curve = 1.0.
- The normal distribution has a high peak and light tails compared to the t-distribution
- The normal distribution is defined by the mean and the variance.

Many variables, such as pregnancy lengths, shoe sizes, foot lengths, and other human physical characteristics exhibit these properties: symmetry indicates that the variable is just as likely to take a value a certain distance below its mean as it is to take a value that same distance above its mean; the bell-shape indicates that values closer to the mean are more likely, and it becomes increasingly unlikely to take values far from the mean in either direction.

The particular shape exhibited by these variables has been studied since the early part of the nineteenth century, when they were first called “normal” as a way of suggesting their depiction of a common, natural pattern.

#### Empirical Rules
As explained by the Empirical Rule, 68% of the area of the normal distribution is within one standard deviation of the mean and approximately 95% of the area is within two standard deviations of the mean:

```{r,echo=FALSE}
# Define variable containing url
url <- "https://s3-us-west-2.amazonaws.com/courses-images-archive-read-only/wp-content/uploads/sites/1298/2015/10/03182600/Figure5_3_1.png"
```

<center><img src="`r url`"></center>



Below, we will use the `dnorm` function to generate the density (or point) probability for a specific value for a normal distribution. This function is very useful for creating a plot of a density function of a distribution. In the list of the random number generator functions all the functions started with an “r”, similarly the density functions for all the distributions all start with a “d”.

#### Point probability for a specific value of a standard normal dist
```{r}
dnorm(-1.96)
```

#### Plotting the density function of a normal distribution: N(2, .25)
```{r}
x <- seq(-5, 4, 0.1)
plot(x, dnorm(x, 2, 0.5), type = "l")
lines(x, dnorm(x, 0, 1), col= "red")
```

It is also possible to calculate p-values using the cumulative distribution functions. For the normal distribution this function is the `pnorm`.

```{r}
# calculating the p-values for the quantiles of a standard normal
1 - pnorm(1.959964)
```

```{r, echo = FALSE}
# (gg)plot and (gg)highlight
# Generate data (see comment below)
library(dplyr)
df <- data.frame(x = seq(-5, 5, length.out = 100)) %>% mutate(y = dnorm(x))
library(ggplot2)
library(gghighlight)
ggplot(df, aes(x, y)) + geom_area(fill = "sky blue") + gghighlight(x > 1.959964) +
theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
```


```{r}
# calculating the p-values for the quantiles of a standard normal
pnorm(1.644854)
```


